---
title: 'CSSS 510: Lab 1'
output:
  beamer_presentation
date: "2017-9-29"
subtitle: Logistics & R Refresher
---

# Logistics

1. **Lab Sessions**: Fri, 4:00-5:20pm in Savery 117
    + Covers application of material from lecture using examples; clarification and extention of lecture material; Q & A for homeworks and lectures
    + Materials will be available on the [**course website**](http://faculty.washington.edu/cadolph/index.php?page=21) 

2. **Office Hours**: Tues, 3:00-4:20pm in Smith 220
    + Available for trouble shooting and specific questions about homework and lecture materials
    
3. **Homeworks**: 5-6 due every 2 weeks or so
    + Ideally, done using $\textsf{R}$ or $\textsf{R}$ Studio with write up in \LaTeX 
    + Using $\textsf{R}$ Studio with $\textsf{R}$ Markdown is an easy way to do this 
    + We will use two of Chris's packages extensively: $\textsf{simcf}$ and $\textsf{tile}$

# Survey Results

\centering
\includegraphics{Rplot.pdf}

# Survey Results

\centering
\includegraphics{latexplot.pdf}

# Survey Results

\centering
\includegraphics{logitplot.pdf}

# Survey Results

\centering
\includegraphics{probitplot.pdf}

# Survey Results

\centering
\includegraphics{mlplot.pdf}


# Logistics

1. When this course is over, you should be able to do the following (and much more):
    + Fit a logistic regression model using both the $\textsf{glm}$ function and "by hand" using $\textsf{optim}$, extract parameters of interest, and interpret these in probabilities
    + Compute predicted probabilities and use simulation to find the confidence intervals of $\hat{\pi}$ across counterfactuals values of $\boldsymbol{x}$ 
    + Use cross-validation to assess the predictive accuracy of several models and also compare these models across a variety of in-sample goodness of fit tests
    + Fit a variety of bounded and unbounded count models that address overdispersion 
    + Use one of several algorithms to impute missing data
$\newline$

# Logistics

2. The course moves fast: you should comfortable doing the following for the homework assignments and project
    + tidying and transforming data
    + importing and exporting data sets 
    + generating plots of your data and results
    + writing basic functions and loops for repeated procedures
$\newline$

* Fortunately, for those of you new to $\textsf{R}$, there are many resources to get you up to speed
    + [Zuur et al. (2009)](https://canvas.uw.edu/courses/1064065/files), Chapter 1-5
    + [Wickham and Groleman (2017)](http://r4ds.had.co.nz/)
    
# R Refresher

## Data Objects 

### Create the following vectors 

1. vector.1 : 1,2,3,4,5,6,6,6,6,6

2. vector.2: 10 randomly drawn numbers from a normal distribution with a mean 10 and a s.d. of 1

3. vector.3: Results of 10 single binomial trials with a probability of 0.4

4. vector.4: Sample 100 observations from a 5-trial binomial distribution with a probability of success of 0.4
$\newline$

# Vectors

```{r}

#Clear memory 
rm(list=ls())

vector.1 <- c(seq(1,5,1), rep(6,5))

vector.2 <- rnorm(10, 10, 1)

#help?
?rnorm

vector.3 <- rbinom(10, 1, 0.4)

vector.4 <- rbinom(100, 5, 0.4)

```

# Vectors

5. Check what type of data vector.2 is

6. Round up vector.2 to two decimal place
$\newline$

# Vectors
\scriptsize
```{r}

is.character(vector.2)
mode(vector.2)

round(vector.2, 2)

```

# Matrices

7. matrix.1: Create 5 by 5 matrix containing all NAs 

8. Assign matrix.1 the row names (a,b,c,d,e) and the column names (1,2,3,4,5)

9. Replace the NAs in the first columne of matrix.1 with Inf
$\newline$

# Matrices 

```{r}

matrix.1<-matrix(NA, nrow=5, ncol=5)

rownames(matrix.1)<-c("a","b","c","d","e")
colnames(matrix.1)<-c(1,2,3,4,5) 

matrix.1[,1]<-Inf

```

# Lists

10. Create a list that contains vector.1, vector.2, and matrix.1

11. Locate vector.2 from the list

# Lists 
\scriptsize
```{r}

list.1 <- list(vector.1, vector.2, vector.3, matrix.1)
names(list.1) <-
  c("vector.1", "vector.2", "vector.3", "matrix.1")

list.1[[2]]
list.1$vector.2

```

# Data Frames 

Data frames are a special type of list in which each row has same length. It is also a matrix like object, yet its elements - unlike elements in a matrix - doesn't have to be of same type. Most of the data we use are in data frames. 

12. Open Lab1data.csv in R

13. Is it a data frame? Is it a matrix?

14. Check the names and summary statistics of the data

15. Remove observations with missing values

16. Plot GDP per capita (on the x-axis) and polity2 (on the y-axis)

17. Create a new variable called "democracy". Assign 0 to countries with negative value or zero polity2 score, and assign 1 to countries with positive score. 

18. Use a loop to do the same recoding
$\newline$

# Data Frames
```{r}

library(foreign)
library(tidyverse)

setwd("/Users/danielyoo/CSSS-POLS-510-MLE/Lab1Slides")

data<-read.csv("Lab1data.csv", header=T)  
```


# Data Frames
```{r}

is.data.frame(data) #Yes!
is.matrix(data) #No
is.character(data$Year)
data$Year<-as.character(data$Year)
```

# Data Frames
```{r}

names(data)
```

# Data Frames
\small
```{r}
summary(data)
```

# Data Frames
\small
```{r}
head(unique(data$country)) # observations on 174 countries
head(tapply(data$country, data$Year, length))
head(tapply(data$Year, data$country, length))
```

# Data Frames
\small
```{r}
data<-na.omit(data) # listwise deletion!!

dim(data)

attach(data)

```


# Data Frames
\tiny
```{r}
plot(polity2, GDP.per.capita.PPP.current.international, ylab="Polity2", xlab="GDP per capita")

```

# Data Frames
```{r}

data$democracy[data$polity2>0]<-1
data$democracy[data$polity2<0|data$polity2==0]<-0
summary(data$democracy)
```


# Data Frames
\small
```{r}
data$democracy.2<-rep(NA, length(data$polity2)) # 1305

for (i in 1:length(data$polity2)) {
    if (data$polity2[i]>0) data$democracy.2[i]<-1
     else data$democracy.2[i]<-0
     }

head(cbind(data$democracy, data$democracy.2))
```

# Data Frames

19. Subset the data frame to show only country name and GDP per capita

20. Rearrange the columns of the data frame ascending by polity score

21. Show only values of GDP per capita for South Africa from 2002 to 2008

22. Create a new variable that takes the first letter of the country and attaches it to the year of observation

23. Find the mean of GDP per capita for each year of observation

# Data Frames
\tiny
```{r}
library(tidyverse)
head(select(data, country, GDP.per.capita.PPP.current.international))
head(data[, c(1,3)])
head(data.frame(data$country, data$GDP.per.capita.PPP.current.international))

```

# Data Frames
\tiny
```{r}
head(arrange(data, polity2))
head(data[order(data$polity2),])
```

# Data Frames
\tiny
```{r}
head(filter(data, country==c("South Africa"), Year>=2002 & Year<=2008))
head(subset(data, data$country==c("South Africa") & data$Year>=2002 & Year<=2008))
```

# Data Frames
\scriptsize
```{r}
head(mutate(data, paste(substring(data$country, 1, 1), data$Year, sep="")))
```

# Data Frames
\scriptsize
```{r}
data%>%
  group_by(Year)%>%
  summarize(mean(GDP.per.capita.PPP.current.international, na.rm=T)
            )
```

